Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-10-11 paper/paper.tex
Fixed some font/symbol stuff

%%
%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[acmsmall,screen=true,
\ifpublic review=false\else,review=true\fi
  ,anonymous=\ifanonymous true\else false\fi]{acmart}
\usepackage{ottalt}
\usepackage{minted}
\usepackage{xspace}
\usepackage[para]{footmisc}
\newcommand{\dotv}[2]{\href{#1}{\texttt{#1}}{\texttt{:#2}}}
\newcommand{\lang}{$\lambda^{\Pi}$\xspace}
\inputott{rules}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%% to allow unicode characters in minted code
\usepackage{newunicodechar}
\newunicodechar{⇒}{$\Rightarrow$}

\usepackage{draft}

@@

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Yiyun Liu}
\orcid{0009-0006-8717-2498}
\affiliation{
  \institution{University of Pennsylvania}
  \city{Philadelphia}
  \country{USA}
}
\email{liuyiyun@seas.upenn.edu}

\author{Stephanie Weirich}
\orcid{0000-0002-6756-9168}
\affiliation{
  \institution{University of Pennsylvania}
  \city{Philadelphia}
  \country{USA}
}
\email{sweirich@seas.upenn.edu}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Proof by logical relations is a powerful technique that has been used
to derive metatheoretic properties of type systems, such as
consistency and parametricity. While there exists a
plethora of introductory materials about logical relation in the
context of simply typed or polymorphic lambda calculus, a streamlined
presentation of proof by logical relation for a dependently typed language
is lacking. In this paper, we present a short
consistency proof for a dependently typed language that contains a
rich set of features, including a countable universe
hierarchy, booleans, and an intensional identity type. We show that
the logical relation can be easily extended to prove the existence of
$\beta\eta$-normal forms.
We have
fully mechanized the consistency proof using the Coq proof assistant
in under 1000 lines of code, with 500 lines of additional code for the
$\beta\eta$-normal form extension.
\end{abstract}

@@

\begin{figure}[h]
\[
\begin{array}{lcll}
% \mathit{Natural\ numbers}\\
% [[i]],[[j]],[[n]] & \in &  [[SNat]] &  \\ \\
\mathit{Terms}\\
[[a]],[[b]],[[c]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[x]]\  |\ [[Void]]
                  & \mbox{universes, variables, empty type} \\
            & |   & [[Pi x : A . B]]\ |\ [[\ x . a]]\ |\ [[a b]]
                  & \mbox{function types, abstractions, applications} \\
            & |   & [[a ~ b : A ]]\ |\  [[refl]]\ |\ [[J c a b p]]
                  & \mbox{equality types, reflexivity proof, J eliminator} \\
            & |   & [[Bool]]\ |\  [[true]]\ |\  [[false]]\ % |\  [[if a b0 b1]]
                  & \mbox{boolean type, true, false} \\
            & |   & [[if a b0 b1]]
                  & \mbox{conditional expression} \\ \\
\end{array}
\]
\[
\begin{array}{lll}
\mathit{Substitutions}      & \qquad\qquad & \mbox{\emph{Typing Contexts}}  \\
[[rho]] \in [[SVar -> STm]] & & [[G]] ::= [[empty]]\ |\ [[G ++ x : A]] 
\end{array}
\]
  \caption{Syntax of \lang}
  \label{fig:syntax}
\end{figure}

@@

\begin{figure}[h]
\begin{minipage}{0.9\textwidth}
\drules[P]{$[[a => b]]$}{Parallel Reduction}{AppAbs, IfTrue, IfFalse, JRefl}
% \drules[PS]{$[[a =>+ b]]$}{Transitive Closure of Parallel Reduction}{Refl, Step}
% \drules[C]{$[[a <=> b]]$}{Convertibility}{Intro}
\end{minipage}
\caption{Parallel reduction ($\beta$-rules only) }
\label{fig:par}
\end{figure}

@@

\begin{figure}[h]
\begin{minipage}{0.9\textwidth}
\drules[Ctx]{$[[ |- G]]$}{Context Well-Formedness}{}
\[ \drule{Ctx-Empty} \qquad \drule[width=3in]{Ctx-Cons} \]
\drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv,
 Void, Bool, True, False}
\[ \drule[width=5in]{T-If}\]
\[ \drule[width=3in]{T-Eq} \qquad \drule{T-Refl} \]
\[ \drule[width=4in]{T-J} \]
\end{minipage}
\caption{Syntactic typing for \lang}
\label{fig:typing}
\end{figure}

@@

\subsection{Definition of the Logical Relation}
\begin{figure}[h]
\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Set, Void, Bool, Eq, Red}
\[ \drule[width=5in]{I-Pi} \]
\caption{Logical relation for \lang}
\label{fig:logrel}
\end{figure}
The logical relation for \lang{}, which takes the form $[[Interp I i A
S]]$, is defined as an inductively generated relation (Figure~\ref{fig:logrel}).
% The logical relation takes the form $[[Interp I i A S]]$.
Metavariables $[[A]]$ and $[[i]]$ stand for terms and natural
numbers respectively, as introduced earlier in
Figure~\ref{fig:syntax}.
% \scw{Many introductory texts define the relation as a recursive function
% over type structure, or step-indices. You use an inductive relation instead, why?}
%\yl{resolved}
% \scw{Can we view this inductive relation as the graph of the partial function
% that is defined recursively over types?} \yl{reolsved}
% \scw{What is this form extensible too? impredicative quantification? recursive
% types? } \yl{resolved? impredicativity is hard}
% \scw{Is it worth observing here that this definition is not over sets of typed
% terms. That it characterizes all terms that look like booleans (i.e. evaluate to
% true or false) or all terms that look like proofs (i.e. evaluate to refl). The
% fact that there is no connection between p and a and b in the I-Eq case is strange
% looking. Need to explain.  }
% \yl{resolved}
The metavariables $[[I]]$ and $[[S]]$ are
sets with the following signatures.
\[
    [[I]] \in [[ { j | j < i  } ->  PowerSet STm ]] \qquad\qquad\qquad
    [[S]] \in [[PowerSet STm]]
\]
The notation $[[PowerSet STm]]$ denotes the powerset of the set of
\lang{} terms.
The function $[[I]]$ is a family of sets of terms indexed by
natural numbers strictly less than the parameter $[[i]]$, which
represents the current universe level.  In \rref{I-Set}, the function
$[[I]]$ is used to define the meaning of
universes that are strictly smaller than the current level $[[i]]$. The
restriction $[[j < i]]$ in \rref{I-Set} ensures the predicativity of the system.

@@

\section{Existence of $\beta$-normal forms}
\label{sec:extension}
In this section, we show how the logical relation from
Section~\ref{sec:logreldep} can be extended to show the existence of $\beta$
normal forms for (open and closed) well-typed terms.  In other words, we prove
that it is possible to repeatedly use the parallel reduction relation to
reduce any term to its unique normal form, where no further (non-identity)
reductions can be applied. This result can be used to show that our type
conversion relation is decidable, as computation of the normal form will
always terminate.

@@

We begin this part with a description of the $\beta$-normal forms
of \lang{}.
%
\begin{figure}[h]
  \[
    \begin{array}{llcl}
       \beta\text{-}\mathit{neutral\ terms} &
      [[e]] & ::= & [[i]]\ |\ [[e f]]\ |\ [[J e f f f]]\ |\ [[if e f
                    f]] \\ \\
      \beta\text{-}\mathit{normal\ terms} &
      [[f]] & ::= & [[e]]\ |\ [[Set i]]\ |\ [[Void]]\ |\ [[Pi x : f . f]]\
                    |\ [[f ~ f : f]]\\
            & & |   & [[\ x . f]]\ |\ [[refl]]\ |\ [[Bool]]\ |\ [[true]]\ |\ [[false]]
    \end{array}
  \]
  \caption{$\beta$-neutral and normal forms}
  \label{fig:nenf}
\end{figure}
%
The syntactic forms $[[e]]$ and $[[f]]$ (Figure~\ref{fig:nenf}) capture the
neutral terms and normal forms with respect to $\beta$-reduction\scw{Why not say parallel reduction here? We can be specific and say that that the only reductions available for these terms are identity reductions and cite
\footnote{\dotv{normalform.v}{nf\_refl}} }.\yl{I always think of
normal form as the specific definition that says the relation can't
step. We can refer to nfrefl but that requires some explanation about
the definition of normal form and the complication that it doesn't
hold in $\eta$ (maybe it's fine if we just don't mention it in the
$\eta$ case)}\scw{A terminal form is a syntactic characterization of terms that don't step 
according to a particular relation. A normal form is a terminal form of a normalizing relation.}
 Instead of the
metavariables $e$ and $f$, we also
use the judgment forms $[[ne a]]$ and $[[nf a]]$ to indicate that there exists
$[[e]]$ or $[[f]]$ such that $[[a = e]]$ or $[[a = f]]$.

@@

\begin{figure}[h]
  \drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Ne, VoidNew, BoolNew, EqNew}
  \caption{Extended logical relation (new and changed rules) }
  \label{fig:logrelopen}
\end{figure}

@@

\section{Existence of $\beta\eta$-normal forms}
\label{sec:betaeta}
\citet{nbeincoq,decagda,martin-lof-a-la-coq}
include the $\eta$ law for functions in their equational theory and
use relational models to justify its validity.
In our system, we can easily incorporate the function $\eta$ law to the equational
theory of \lang{} by adding the following parallel reduction rule.
\begin{center}
  \drule[width=2.5in]{P-AbsEta}
\end{center}
In this section, we show how we easily extend the existence of $\beta$-normal
forms from Section~\ref{sec:extension} to the existence of
$\beta\eta$-normal forms after this addition.

@@

% \scw{Is this specific to $\eta$? Or is this also true for $\beta$? Can we move this
% sentence elsewhere? I'd like to use the previous sentence as a conclusion for
% this section.}
% Finally, due to technical
% details, our mechanized proof about the existence of $\beta\eta$ form does not
% immediately induce a decision procedure for type conversion in
% \lang{}. We discuss this issue and its workaround in
% Section~\ref{sec:conversionalgo}.
\section{Mechanization}
\label{sec:logrelmech}
\begin{figure}[h]
  \begin{minipage}{0.9\textwidth}
  \begin{tabular}{ |l |  c  | c | c| }
    \hline
    & Consistency & Normalization & Syntactic metatheory \\
    \hline
    Syntactic typing         &  83 & =   & = \\
    Untyped reduction        & 344 & =   & = \\
    Neutral and normal forms & -   & 273 & - \\
    Logical relation         & 338 & 430 & - \\
    Semantic soundness       & 192 & 211 & - \\
    Syntactic soundness      & -   & -   & 629 \\
    \hline
    Total                    & 957 & 1341 & 1056 \\
    \hline
  \end{tabular}
  \end{minipage}
  \caption{Nonblank, noncomment lines of code of the Coq Development. The marker = indicates that the line count is the same as the column to the
left. The marker - indicates the file does not
contribute to the total.}
  \label{fig:linecount}
\end{figure}

@@

For other automation tasks that are not specific to binding, we use
the powerful \texttt{sauto} tactic provided by CoqHammer to write
short and declarative proofs. For example, here is a one-line proof of
the triangle property about parallel reduction, from which the diamond
property (Lemma~\ref{lemma:pardiamond}) follows as a corollary.
The triangle property states
that if $[[a => b]]$, then $[[b]]\Rightarrow [[a]]^*$, where $[[a]]^*$
is the Takahashi translation~\citep{takahashi-parallel-reduction}
which roughly corresponds to simultaneous reduction of the redexes in
$[[a]]$, excluding the new redexes that appear as a result of
reduction.
\begin{minted}{coq}
Lemma Par_triangle a : forall b, (a ⇒ b) -> (b ⇒ tstar a).
Proof.
  apply tstar_ind; hauto lq:on inv:Par use:Par_refl,Par_cong ctrs:Par.
Qed.
\end{minted}
In prose, the triangle property can be proven by induction over the graph of
\mintinline{coq}{tstar a}, the Takahashi translation. Options \texttt{inv:Par}
and \texttt{ctrs:Par} say that the proof involves inverting and constructing
of the derivations of parallel reduction. The option
\texttt{use:Par\_refl,par\_cong} allows the automation tactic to use the
reflexivity and congruence properties of parallel reduction as lemmas.

@@

\begin{figure}[h]

@@

    \hline
    \lang{} (this work) & $\mathbb{N}$ & Id, Bool & U & {\boxedsymbols ✓} & 1 
    & Consistency and normalization \\
    % \hline
    $\lambda^\theta$ & 0 & Id, Nat & U & {\boxedsymbols ✗} & 1 & Consistency \\
    Core Nuprl & $\mathbb{N}$ & W-Types & E & {\boxedsymbols ✓} & 2 & Consistency \\
    NBE-in-Coq & 1 & Nat & T & {\boxedsymbols ✗} & 2 & Correctness of NBE \\
    $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat, $\Sigma$ & T & {\boxedsymbols ✓} & 2 & Decidability of conversion \\
    MLTT-\'a-la-Coq & 1 & Id, Nat, $\Sigma$ & T & {\boxedsymbols ✓} & 2 & Decidibility of type checking \\
    \hline
  \end{tabular}

  \begin{tabular}{ll}
    \\
  \underline{U}niverses: & Countable ($\mathbb{N}$), Zero (0), One (1) \\
  \underline{Ind}uctives: & Identity types (Id), Natural numbers (Nat), $\Sigma$-types ($\Sigma$), W-types \\
  \underline{C}onversion:& Untyped (U), Typed (T),  Extensional (E) \\
  \underline{L}arge \underline{E}liminations:& Included ({\boxedsymbols ✓}), not included ({\boxedsymbols ✗}) \\
  \underline{A}rity of interpretation: & Sets of terms (1), Relations between terms (2) \\
  \\
  $\lambda^\theta$  & \citet{casinghino:combining-proofs-programs} (logical fragment only) \\
  Core Nuprl &\citet{anand2014towards} \\
  $\lambda^{\Pi U\mathbb{N}}$ &\citet{decagda} \\
  NBE-in-Coq& \citet{nbeincoq} \\
  MLTT-\'a-la-Coq &\citet{martin-lof-a-la-coq} \\
  \end{tabular}

>>>>

%%
%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall')
%%
%% IMPORTANT NOTICE:
%%
%% For the copyright see the source file.
%%
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%%
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%%
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[\ifpublic nolinenum\else\fi]{jfp}

\let\Bbbk\relax
\usepackage{ottalt}
\usepackage[outputdir=latex.out]{minted}
\usepackage{xspace}
\usepackage{stmaryrd}
\usepackage{bbding}
\usepackage{amsthm,amssymb}
\usepackage{dejavu}
\usepackage[para]{footmisc}
\newcommand{\dotv}[2]{\href{#1}{\texttt{#1}}{\texttt{:#2}}}
\newcommand{\lang}{$\lambda^{\Pi}$\xspace}
\inputott{rules}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

\let\draftnote\relax % this is already defined in JFP
\usepackage{draft}

@@

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}

\DeclareTextCommand{\nobreakspace}{TU}{\leavevmode\nobreak\ }

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title[T]{Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Yiyun Liu}
% \orcid{0009-0006-8717-2498}
\affiliation{University of Pennsylvania (\email{liuyiyun@seas.upenn.edu})}

\author{Stephanie Weirich}
% \orcid{0000-0002-6756-9168}
\affiliation{University of Pennsylvania (\email{sweirich@seas.upenn.edu})}

\author{Jonathan Chan}
\affiliation{University of Pennsylvania (\email{jcxz@seas.upenn.edu})}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Proof by logical relations is a powerful technique that has been used
to derive metatheoretic properties of type systems, such as
consistency and parametricity. While there exists a
plethora of introductory materials about logical relations in the
context of simply typed or polymorphic lambda calculi, a streamlined
presentation of proof by logical relation for a dependently typed language
is lacking. In this paper, we present a short
consistency proof for a dependently typed language that contains a
rich set of features, including a countable universe
hierarchy, booleans, and an intensional identity type. We show that
the logical relation can be easily extended to prove the existence of
$\beta\eta$-normal forms.
We have
fully mechanized the consistency proof using the Coq proof assistant
in under 1000 lines of code, with 500 lines of additional code for the
$\beta\eta$-normal form extension.
\end{abstract}

@@

\begin{figure}[H]
\[
\begin{array}{lcll}
% \mathit{Natural\ numbers}\\
% [[i]],[[j]],[[n]] & \in &  [[SNat]] &  \\ \\
\mathit{Terms}\\
[[a]],[[b]],[[c]],[[p]],[[A]],[[B]] & ::= & [[Set i]]\ |\ [[x]]\  |\ [[Void]]
                  & \mbox{universes, variables, empty type} \\
            & |   & [[Pi x : A . B]]\ |\ [[\ x . a]]\ |\ [[a b]]
                  & \mbox{function types, abstractions, applications} \\
            & |   & [[a ~ b : A ]]\ |\  [[refl]]\ |\ [[J c a b p]]
                  & \mbox{equality types, reflexivity proof, J eliminator} \\
            & |   & [[Bool]]\ |\  [[true]]\ |\  [[false]]\ % |\  [[if a b0 b1]]
                  & \mbox{boolean type, true, false} \\
            & |   & [[if a b0 b1]]
                  & \mbox{conditional expression} \\ \\
\end{array}
\]
\[
\begin{array}{lll}
\mathit{Substitutions}      & \qquad\qquad & \mbox{\emph{Typing Contexts}}  \\
[[rho]] \in [[SVar -> STm]] & & [[G]] ::= [[empty]]\ |\ [[G ++ x : A]] 
\end{array}
\]
  \caption{Syntax of \lang}
  \label{fig:syntax}
\end{figure}

@@

\begin{figure}[H]
\begin{minipage}{0.9\textwidth}
\drules[P]{$[[a => b]]$}{Parallel Reduction}{AppAbs, IfTrue, IfFalse, JRefl}
% \drules[PS]{$[[a =>+ b]]$}{Transitive Closure of Parallel Reduction}{Refl, Step}
% \drules[C]{$[[a <=> b]]$}{Convertibility}{Intro}
\end{minipage}
\caption{Parallel reduction ($\beta$-rules only) }
\label{fig:par}
\end{figure}

@@

\begin{figure}[H]
\begin{minipage}{0.9\textwidth}
\drules[Ctx]{$[[ |- G]]$}{Context Well-Formedness}{}
\[ \drule{Ctx-Empty} \qquad \drule[width=3in]{Ctx-Cons} \]
\drules[T]{$[[G |-  a : A]]$}{Typing}{Var, Set, Pi, Abs, App, Conv,
 Void, Bool, True, False}
\[ \drule[width=5in]{T-If}\]
\[ \drule[width=3in]{T-Eq} \qquad \drule{T-Refl} \]
\[ \drule[width=4in]{T-J} \]
\end{minipage}
\caption{Syntactic typing for \lang}
\label{fig:typing}
\end{figure}

@@

\subsection{Definition of the Logical Relation}
\begin{figure}[H]
\drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Set, Void, Bool, Eq, Red}
\[ \drule[width=5in]{I-Pi} \]
\caption{Logical relation for \lang}
\label{fig:logrel}
\end{figure}
The logical relation for \lang{}, which takes the form $[[Interp I i A
S]]$, is defined as an inductively generated relation (Figure~\ref{fig:logrel}).
% The logical relation takes the form $[[Interp I i A S]]$.
Metavariables $[[A]]$ and $[[i]]$ stand for terms and natural
numbers respectively, as introduced earlier in
Figure~\ref{fig:syntax}.
% \scw{Many introductory texts define the relation as a recursive function
% over type structure, or step-indices. You use an inductive relation instead, why?}
%\yl{resolved}
% \scw{Can we view this inductive relation as the graph of the partial function
% that is defined recursively over types?} \yl{reolsved}
% \scw{What is this form extensible too? impredicative quantification? recursive
% types? } \yl{resolved? impredicativity is hard}
% \scw{Is it worth observing here that this definition is not over sets of typed
% terms. That it characterizes all terms that look like booleans (i.e. evaluate to
% true or false) or all terms that look like proofs (i.e. evaluate to refl). The
% fact that there is no connection between p and a and b in the I-Eq case is strange
% looking. Need to explain.  }
% \yl{resolved}
The metavariables $[[I]]$ and $[[S]]$ are
sets with the following signatures.
\[
    [[I]] \in [[ { j | j < i  } ->  PowerSet STm ]] \qquad\qquad\qquad
    [[S]] \in [[PowerSet STm]]
\]
The notation $[[PowerSet STm]]$ denotes the powerset of the set of
\lang{} terms.
The function $[[I]]$ is a family of sets of terms indexed by
natural numbers strictly less than the parameter $[[i]]$, which
represents the current universe level.  In \rref{I-Set}, the function
$[[I]]$ is used to define the meaning of
universes that are strictly smaller than the current level $[[i]]$. The
restriction $[[j < i]]$ in \rref{I-Set} ensures the predicativity of the system.

@@

\section{Existence of \texorpdfstring{$\beta$}{beta}-normal forms}
\label{sec:extension}
In this section, we show how the logical relation from
Section~\ref{sec:logreldep} can be extended to show the existence of $\beta$
normal forms for (open and closed) well-typed terms.  In other words, we prove
that it is possible to repeatedly use the parallel reduction relation to
reduce any term to its unique normal form, where no further (non-identity)
reductions can be applied. This result can be used to show that our type
conversion relation is decidable.% , as computation of the normal form will
% always terminate.\yl{dependent on reduction strategy}

@@

We begin this part with a description of the $\beta$-normal forms
of \lang{}.
%
\begin{figure}[H]
  \[
    \begin{array}{llcl}
       \beta\text{-}\mathit{neutral\ terms} &
      [[e]] & ::= & [[x]]\ |\ [[e f]]\ |\ [[J e f f f]]\ |\ [[if e f
                    f]] \\ \\
      \beta\text{-}\mathit{normal\ terms} &
      [[f]] & ::= & [[e]]\ |\ [[Set i]]\ |\ [[Void]]\ |\ [[Pi x : f . f]]\
                    |\ [[f ~ f : f]]\\
            & & |   & [[\ x . f]]\ |\ [[refl]]\ |\ [[Bool]]\ |\ [[true]]\ |\ [[false]]
    \end{array}
  \]
  \caption{$\beta$-neutral and normal forms}
  \label{fig:nenf}
\end{figure}
%
The syntactic forms $[[e]]$ and $[[f]]$ (Figure~\ref{fig:nenf}) capture the
neutral terms and normal forms with respect to $\beta$-reduction\scw{Why not say parallel reduction here? We can be specific and say that that the only reductions available for these terms are identity reductions and cite
\footnote{\dotv{normalform.v}{nf\_refl}} }.\yl{I always think of
normal form as the specific definition that says the relation can't
step. We can refer to nfrefl but that requires some explanation about
the definition of normal form and the complication that it doesn't
hold in $\eta$ (maybe it's fine if we just don't mention it in the
$\eta$ case)}\scw{A terminal form is a syntactic characterization of terms that don't step 
according to a particular relation. A normal form is a terminal form of a normalizing relation.}
 Instead of the
metavariables $e$ and $f$, we also
use the judgment forms $[[ne a]]$ and $[[nf a]]$ to indicate that there exists
$[[e]]$ or $[[f]]$ such that $[[a = e]]$ or $[[a = f]]$.

@@

\begin{figure}[H]
  \drules[I]{$[[Interp I i A S]]$}{Logical Relation}{Ne, VoidNew, BoolNew, EqNew}
  \caption{Extended logical relation (new and changed rules) }
  \label{fig:logrelopen}
\end{figure}

@@

\section{Existence of \texorpdfstring{$\beta\eta$-}{beta-eta--}normal forms}
\label{sec:betaeta}
\citet{nbeincoq,decagda,martin-lof-a-la-coq}
include the $\eta$ law for functions in their equational theory and
use relational models to justify its validity.
In our system, we can easily incorporate the function $\eta$ law to the equational
theory of \lang{} by adding the following parallel reduction rule.
\begin{center}
  \drule[width=2.5in]{P-AbsEta}
\end{center}
In this section, we show how we easily extend the existence of $\beta$-normal
forms from Section~\ref{sec:extension} to the existence of
$\beta\eta$-normal forms after this addition.

@@

% \scw{Is this specific to $\eta$? Or is this also true for $\beta$? Can we move this
% sentence elsewhere? I'd like to use the previous sentence as a conclusion for
% this section.}
% Finally, due to technical
% details, our mechanized proof about the existence of $\beta\eta$ form does not
% immediately induce a decision procedure for type conversion in
% \lang{}. We discuss this issue and its workaround in
% Section~\ref{sec:conversionalgo}.
\section{Mechanization}
\label{sec:logrelmech}
\begin{figure}[H]
  \begin{minipage}{0.9\textwidth}
  \begin{tabular}{ |l |  c  | c | c| }
    \hline
    & Consistency & Normalization & Syntactic metatheory \\
    \hline
    Syntactic typing         &  83 & =   & = \\
    Untyped reduction        & 344 & =   & = \\
    Neutral and normal forms & -   & 273 & - \\
    Logical relation         & 338 & 430 & - \\
    Semantic soundness       & 192 & 211 & - \\
    Syntactic soundness      & -   & -   & 629 \\
    \hline
    Total                    & 957 & 1341 & 1056 \\
    \hline
  \end{tabular}
  \end{minipage}
  \caption{Nonblank, noncomment lines of code of the Coq Development. The marker = indicates that the line count is the same as the column to the
left. The marker - indicates the file does not
contribute to the total.}
  \label{fig:linecount}
\end{figure}

@@

For other automation tasks that are not specific to binding, we use
the powerful \texttt{sauto} tactic provided by CoqHammer to write
short and declarative proofs. For example, here is a one-line proof of
the triangle property about parallel reduction, from which the diamond
property (Lemma~\ref{lemma:pardiamond}) follows as a corollary.
The triangle property states
that if $[[a => b]]$, then $[[b]]\Rightarrow [[a]]^*$, where $[[a]]^*$
is the Takahashi translation~\citep{takahashi-parallel-reduction}
which roughly corresponds to simultaneous reduction of the redexes in
$[[a]]$, excluding the new redexes that appear as a result of
reduction.
\begin{minted}[escapeinside=||,mathescape=true]{coq}
Lemma Par_triangle a : forall b, (a |$\Rightarrow$| b) -> (b |$\Rightarrow$| tstar a).
Proof.
  apply tstar_ind; hauto lq:on inv:Par use:Par_refl,Par_cong ctrs:Par.
Qed.
\end{minted}
In prose, the triangle property can be proven by induction over the graph of
\mintinline{coq}{tstar a}, the Takahashi translation. Options \texttt{inv:Par}
and \texttt{ctrs:Par} say that the proof involves inverting and constructing
of the derivations of parallel reduction. The option
\texttt{use:Par\_refl,Par\_cong} allows the automation tactic to use the
reflexivity and congruence properties of parallel reduction as lemmas.

@@

\begin{figure}[H]

@@

    \hline
    \lang{} (this work) & $\mathbb{N}$ & Id, Bool & U & $\checkmark$ & 1 
    & Consistency and normalization \\
    % \hline
    $\lambda^\theta$ & 0 & Id, Nat & U & $\XSolidBrush$ & 1 & Consistency \\
    Core Nuprl & $\mathbb{N}$ & W-Types & E & $\checkmark$ & 2 & Consistency \\
    NBE-in-Coq & 1 & Nat & T & $\XSolidBrush$ & 2 & Correctness of NBE \\
    $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat, $\Sigma$ & T & $\checkmark$ & 2 & Decidability of conversion \\
    MLTT-\'a-la-Coq & 1 & Id, Nat, $\Sigma$ & T & $\checkmark$ & 2 & Decidibility of type checking \\
    \hline
  \end{tabular}

  \begin{tabular}{ll}
    \\
  \underline{U}niverses: & Countable ($\mathbb{N}$), Zero (0), One (1) \\
  \underline{Ind}uctives: & Identity types (Id), Natural numbers (Nat), $\Sigma$-types ($\Sigma$), W-types \\
  \underline{C}onversion:& Untyped (U), Typed (T),  Extensional (E) \\
  \underline{L}arge \underline{E}liminations:& Included ($\checkmark$), not included ($\XSolidBrush$) \\
  \underline{A}rity of interpretation: & Sets of terms (1), Relations between terms (2) \\
  \\
  $\lambda^\theta$  & \citet{casinghino:combining-proofs-programs} (logical fragment only) \\
  Core Nuprl &\citet{anand2014towards} \\
  $\lambda^{\Pi U\mathbb{N}}$ &\citet{decagda} \\
  NBE-in-Coq& \citet{nbeincoq} \\
  MLTT-\'a-la-Coq &\citet{martin-lof-a-la-coq} \\
  \end{tabular}
